A Gentle Introduction to Compressed Sensing

### 2017-01-14 compressed_sensing/cs.tex
cs upd

\usepackage{latexsym,amssymb,amsthm}
\usepackage{amsfonts}
%\usepackage[T2A]{fontenc}
\usepackage{amsmath}
\usepackage{url}

@@

\section{Introduction}
In the last decade the humanity has come to the point when it produces much more information than it can store and analyse.
There is a huge exposion in the variety of sensors and their dimensionality in many areas: video, audio, scientific, medical research.
Because of that we face the problem of \textbf{data deluge} and naturally come to the necessity of compressing data.

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=.75\textwidth]{figures/CS_sample_compress_paradigm.pdf}
    \end{center}
    \caption{
        \label{fig:CSSampleCompressParadigm}
        ``Sample-Then-Compress Paradigm''.
        Sample could be a digital camera, a microphone, \ldots. Compression strategy is signal-dependent.
    }
\end{figure}
Let us consider the so-called ``Sample-Then-Compress Paradigm'' (Fig~\ref{fig:CSSampleCompressParadigm}).
We want to transmit a signal (it could be continuous), but are unable to do it with a whole signal.
Thus, we sample it $n$ times.
The transmission channel we use reduces the ``dimension'' of the sampled signal to $S \ll n$.
Afterwards, the received compressed signal is decompressed to previous dimension which is equal to $n$.
In this scenario $S$ is signal-nature-specific.
First, with $S \ll n$ the process seems to be somewhat wasteful.
Secondly, every sampling (measurement) could be exorbitant.
Thus, we want to reduce $n$ as much as it is possible.

The idea of Compressed Sensing is that the signal at Fig.~\ref{fig:CSSampleCompressParadigm} that we would like to transmit/acquire is actually sparse (or ``approximately sparse'') signal in a certain ``basis''.
Suppose that this signal is an $m$-dimensional vector.
Then we want to find a linear basis in $\mathbb R^m$ for this vector to be $S$-sparse.
The challenge of the problem is that we do not observe the signal explicitly (in any basis).
We can make sampling (measurements) that are a certain transformation of the signal (for example, linear).
Speaking non-formally, our aim is to reconstruct the signal making as few measurements as possible (to reduce $n$ at Fig.~\ref{fig:CSSampleCompressParadigm}).
Without the assumption of the existence of the basis in which the original signal is sparse the problem would have a trivial solution: we would generally need to make as many measurements as is the dimension of a signal.

@@

\section{An example of Compressed Sensing sparsity}
\label{sec:motivation}
The main idea of the Compressed Sensing is that a signal vector of our interest is a \textbf{sparse} vector when represented in \textbf{some basis}.
Technically speaking it could be not sparsity in a strict sense (when many components are zeros), but the approximately sparsity (when many components are close to zero).
There are lots of non-formal examples demonstrating this idea, however, many of them are very difficult to formalize.
We will stick to a very simple example that makes a connection with the time series theory.

Let us consider a time series $x = \{x_n\}_{n=1}^N$.
The time series $x$ itself is very rarely a sparse vector, i.e. one would not rely on the assumption that $x_n = 0$ for many $n$-s.
However, what could be assumed is that the periodogram has only a few peaks.
Indeed, let us suppose for all $n \in 1:N$
$$ x_n = \sum_{i=1}^S \sin\left(2 \pi w_i \frac{n}{N}\right), $$
where $w_i \in [0, 0.5]$ and $w_i \neq w_j$ for $i \neq j$.
We will suppose that the length $N$ is multiple of the period of $x$.
Then the periodogram will have peaks only at $w_i$.
In this sense the \textbf{Fourier basis} is the basis in which $x$ is $S$-\textbf{sparse}.

@@

We fix $m \in \mathbb N$ and will denote the signal $\beta^*$ and will suppose that $\beta^* \in \mathbb R^m$.
We suppose that $\beta^*$ is (strictly) $S$-sparse (has only $S$ non-zero components).
This will be denoted as $\|\beta^*\|_0 = S$.
The $S$ itself is latent (not observed).
We also fix the set of design transformation matrices $\mathbf X_n \in \mathcal M_{n \times m}(\mathbb R)$ for all $n \in \mathbb N$.
The ``measurements'' are denoted $y_n \in \mathbb R^n$ for all $n \in \mathbb N$.
The formal connection between $\beta^*$ and $y_n$ is made through the matrix $\mathbf X_n$ for all $n \in \mathbb N$ and will be explicitly stated.

Let us note one more time that we are aware of $X_n$ for all $n$ and observe $y_n$, but do not observe $\beta^*$ and want to find it with the smallest $n$ possible.

\subsection{Noiseless case}
Let us suppose that $y_n = X_n \beta^*$ for all $n \in \mathbb N$.
The problem is to obtain $\beta^*$ as the solution of the linear system $y_n = \mathbf X_n \beta$ (the variable $\beta$).
The illustration can be seen at Fig.~\ref{fig:NoiseLessProblemStatement}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=.75\textwidth]{figures/problem_statement_2.pdf}
    \end{center}
    \caption{
        \label{fig:NoiseLessProblemStatement}
        Noiseless problem statement.
    }
\end{figure}

@@

\begin{definition}
    \label{def:rip}
    A matrix $\mathbf A \in \mathcal M_{n, m}(\mathbb R)$ is said to satisfy the \textbf{restricted isometry property} of order $S$ with constant $\delta_S$ for $\delta_S \in (0, 1)$ if
    $$ (1 - \delta_S) \|x\|_2^2 \leqslant \|\mathbf A x\|_2^2 \leqslant (1 + \delta_S) \|x\|_2^2 $$
    for all $S$-sparse $x \in \mathbb R^m$.

@@

\begin{theorem}
    Let $\mathbf X_n \in \mathcal M_{n, m} (\mathbb R)$ be $\textrm{RIP}(S, \delta)$ where $S \leq m / 2$ and $\delta \in (0, 1)$. Then for
    $$ n \geqslant C_\delta S \log(m / S), $$
    where $C_\delta < 1$ is a constant and depends only on $\delta$:
    $$ \hat \beta_n = \beta. $$
\end{theorem}
Proof can be found in Theorem 3.5: \url{mdav.ece.gatech.edu/publications/phdthesis-2010.pdf}.

It is an NP-problem to check whether the matrix satisfies RIP.
However, it is easy to simulate a matrix that satisfies RIP with a high probability.

\begin{theorem}
    Fix $\delta \in (0, 1)$.
    Let $\mathbf X_n \in \mathcal M_{n, m} (\mathbb R)$ be a random matrix whose components are independent and have $\mathcal N(0, 1/n^2)$ distribution.
    If for some $C_1 > 0$
    $$ n \geqslant C_1 \log(m / S), $$
    then 
    $$ \mathbb P(\mathbf X_n \in \textrm{RIP}(S, \delta)) \geq 1 - 2 e^{-C_2 n}, $$
    where $C_2 = \delta^2/2C_1 - \ln (42 e / \delta) / C_1$.
\end{theorem}
Proof can be found in Theorem 4.3: \url{mdav.ece.gatech.edu/publications/phdthesis-2010.pdf}.

Let us note that the RIP quality is not the only condition that allows to formulate such results.
There are plenty of them, see Section 3.1.3: \url{https://sites.google.com/site/igorcarron2/cs}

>>>>

\usepackage{latexsym,amssymb,amsthm}
\usepackage{amsfonts}
%\usepackage[T2A]{fontenc}
\usepackage{amsmath}
\usepackage{url}
\usepackage{pgfplots}

@@

\section{Motivational introduction}
In the last decade the humanity has come to the point when it produces much more information than it can store and analyse.
There is a huge exposion in the variety of sensors and their dimensionality in many areas: video, audio, scientific, medical research.
Because of that we face the problem of \textbf{data deluge} and naturally come to the necessity of compressing data.

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=.75\textwidth]{figures/CS_sample_compress_paradigm.pdf}
    \end{center}
    \caption{
        \label{fig:CSSampleCompressParadigm}
        ``Sample-Then-Compress Paradigm''.
        Sample could be a digital camera, a microphone,~\ldots\, Compression strategy is signal-dependent.
    }
\end{figure}
Let us consider the so-called ``Sample-Then-Compress Paradigm'' (Fig~\ref{fig:CSSampleCompressParadigm}).
We want to transmit a signal (it could be continuous, it does not matter at this point), but we are unable to do it.
Instead we sample it $n$ times (we will say that the dimension of the sample is $n$).
The transmission channel we reduce the dimension of the sampled signal to $S \ll n$.
Afterwards, the received compressed signal is decompressed to previous dimension $n$.
In this scenario $S$ depends on the nature of the signal.
First, with $S \ll n$ the process seems to be somewhat wasteful: why do we actually need to do $n$ samples if we anyway compress in to the dimension of $S$?!
Secondly, every measurement of total $n$ in a sample could be exorbitant.
Thus, we want to reduce $n$ as much as it is possible.

This example is not accurate as we will see, as follows.
However, it demonstrates why people were concerned about Compressed Sensing in the first place.

\section{Introduction to the Compressed Sensing idea}
The idea of Compressed Sensing is that the signal at Fig.~\ref{fig:CSSampleCompressParadigm} that we would like to transmit/acquire is actually sparse (or ``approximately sparse'') signal in a certain ``basis''.
Suppose that this signal is an $m$-dimensional vector, where $m$ is very large.
Then we want to find a linear basis in $\mathbb R^m$ for this vector to be $S$-sparse (i.e. to have only $S$ nonzero components).
The challenge of the problem is that we do not observe the whole signal.\footnote{
    Suppose that we do.
    Then we can use it as an orth in a basis and the signal is $1$-sparse.
    However, we cannot observe the whole signal because $m$ is very large.
    Take a look at Fig.\ref{fig:CSSampleCompressParadigm}.
    If the signal is an $m$-dimensional vector, then to find the basis we would need to take $n := m$ samples.
    But as we have already discussed we want to reduce $n$ as much as it is possible.
    Because of that we want to find a basis of interest without observing the whole signal.
}

Let us suppose that we now deal with a $S$-sparse signal (we still do not observe it, but know that it is $S$-sparse).
Our aim is to reconstruct the signal (think about decompressing at Fig.\ref{fig:CSSampleCompressParadigm}).
We can make measurements that are a certain transformation of the signal (for example, linear).
We want to make as few measurements as possible (to reduce $n$ at Fig.~\ref{fig:CSSampleCompressParadigm}).
Without the assumption of us possessing the basis in which the original signal is sparse the problem would have a trivial solution: we would generally need to make as many measurements as is the dimension of a signal.\footnote{Think it through!}

@@

\section{An example of Compressed Sensing sparsity}
\label{sec:motivation}
The main idea of the Compressed Sensing is that a signal vector of our interest is a \textbf{sparse} vector when represented in \textbf{some basis}.
Technically speaking it could be approximately sparse when many components are close to zero.
There are lots of non-formal examples demonstrating this idea, however, many of them are very difficult to formalize.
We will stick to a very simple example that makes a connection with the time series theory.

Let us consider time series $x = \{x_n\}_{n=1}^N$.
The time series $x$ itself is very rarely a sparse vector, i.e. one would not rely on the assumption that $x_n = 0$ for many $n$-s.
However, what could be assumed is that the periodogram has only a few peaks.
Indeed, let us suppose for all $n \in 1:N$
$$ x_n = \sum_{i=1}^S \sin\left(2 \pi w_i \frac{n}{N}\right), $$
where $w_i \in [0, 0.5]$ and $w_i \neq w_j$ for $i \neq j$.
We will suppose that the length $N$ is multiple of the period of $x$.
Then the periodogram will have peaks only at $w_i$.
In this sense the \textbf{Fourier basis} is the basis in which $x$ is $S$-\textbf{sparse}.
Note, that we do not need to know the whole series $x$ to be sure that we possess a basis in which the signal is $S$-sparse.

@@

We fix $m \in \mathbb N$. The signal is denoted $\beta^* \in \mathbb R^m$.
We suppose that $\beta^*$ is (strictly) $S$-sparse (has only $S$ non-zero components).
This will be denoted as $\|\beta^*\|_0 = S$.
The $S$ itself is latent (not observed).
We also fix the set of design transformation matrices $\mathbf X_n \in \mathcal M_{n \times m}(\mathbb R)$ for all $n \in \mathbb N$.
The ``measurements'' are denoted $y_n \in \mathbb R^n$ for all $n \in \mathbb N$.
The formal connection between $\beta^*$ and $y_n$ is made through the matrix $\mathbf X_n$ for all $n \in \mathbb N$ and will be explicitly stated.

Let us note one more time that we are aware of $\mathbf X_n$ for all $n$ and observe $y_n$, but do not observe $\beta^*$ and want to find it with the smallest $n$ possible.

\subsection{Noiseless case}
Let us suppose that $y_n = \mathbf X_n \beta^*$ for all $n \in \mathbb N$.
The problem is to obtain $\beta^*$ as the solution of the linear system $y_n = \mathbf X_n \beta$ (the variable $\beta$).
The illustration can be seen at Fig.~\ref{fig:NoiseLessProblemStatement}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=.75\textwidth]{figures/problem_statement_2.pdf}
    \end{center}
    \caption{
        \label{fig:NoiseLessProblemStatement}
        Noiseless problem statement.
    }
\end{figure}

@@

\begin{definition}
    \label{def:rip}
    A matrix $\mathbf A \in \mathcal M_{n, m}(\mathbb R)$ is said to satisfy the \textbf{restricted isometry property} (RIP) of order $S$ with constant $\delta_S$ for $\delta_S \in (0, 1)$ if
    $$ (1 - \delta_S) \|x\|_2^2 \leqslant \|\mathbf A x\|_2^2 \leqslant (1 + \delta_S) \|x\|_2^2 $$
    for all $S$-sparse $x \in \mathbb R^m$.

@@

\begin{theorem}
    \label{theorem:CSforRIP}
    Let $\mathbf X_n \in \mathcal M_{n, m} (\mathbb R)$ be $\textrm{RIP}(S, \delta)$ where $S \leq m / 2$ and $\delta \in (0, 1)$. Then for
    $$ n \geqslant C_\delta S \ln(m / S), $$
    where $C_\delta < 1$ is a constant and depends only on $\delta$:
    $$ \hat \beta_n = \beta. $$
\end{theorem}
Proof can be found in Theorem 3.5: \url{mdav.ece.gatech.edu/publications/phdthesis-2010.pdf}.
Let us comment on the result of the theorem.
Suppose, that $S$ is very small comparatively to $m$, then
$$ C_\delta S \ln(m / S) = C_\delta S \ln m - C_\delta S \ln S \approx const \cdot \ln m. $$
It means that we only need to make a number of samples proportional to $\ln m$ instead of $m$ when the assumption about the sparsity is not made.
If $S = m / 2$ then
$$ C_\delta S \ln(m / S) = (C_\delta m \ln 2) / 2 \approx const \cdot m. $$
It means that we need to make a number of samples proportional to $m$.
Consequently, the number of measurements is bounded between $\log m$ and $m$.

The problem of checking the RIP condition for a given matrix is NP-hard.
However, it is easy to simulate a matrix that satisfies RIP with a high probability.

\begin{theorem}
    Fix $\delta \in (0, 1)$.
    Let $\mathbf X_n \in \mathcal M_{n, m} (\mathbb R)$ be a random matrix whose components are independent and have $\mathcal N(0, 1/n^2)$ distribution.
    Fix $C_1 > 0$ such as
    $$ n \geqslant C_1 \ln(m / S), $$
    then 
    $$ \mathbb P(\mathbf X_n \in \textrm{RIP}(S, \delta)) \geq 1 - 2 e^{-C_2 n}, $$
    where $C_2 = \delta^2/2C_1 - \ln (42 e / \delta) / C_1$.
\end{theorem}
Proof can be found in Theorem 4.3: \url{mdav.ece.gatech.edu/publications/phdthesis-2010.pdf}.

As we want to make the probability as high as possible, to produce RIP matrix we want $\exp(-C_2n)$ as small as possible.
It means that $C_2 n$ as large as possible.
Let's take $C_1$: $n = C_1 \ln(m / S)$:
$$C_2 n = C_1 C_2 \ln(m / S) = \ln(m / S) [ \delta^2 / 2 - \ln(42 e / \delta) ] = \ln(m/S) [ \delta^2 / 2 + \ln \delta + const ]. $$
So the probability we are interested in depends on $\delta$: the higher $\delta$ is --- the higher the probability is (with fixed $m$).
Consider an extreme example of $\delta = 1$:
$$ \mathbb P(\mathbf X_n \in \textrm{RIP}(S, \delta)) \geq 1 - 2 e^{-C_2 n} = 1 - 2 e^{ \ln{m/S}(0.5 + \ln(42))} \approx 1 - 2 e^{4.2 \ln(m/S)}. $$
WTF???

Let us note that the RIP quality is not the only condition that allows to formulate results similar to the theorem \ref{theorem:CSforRIP}.
There are plenty of them, see Section 3.1.3: \url{https://sites.google.com/site/igorcarron2/cs}.
RIP is somewhat the most traditional.
